---
title: "Example: MNIST Digits"
author: "Kisung You"
format:
  gfm:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    dpi: 400
---

### Introduction

This notebook reproduces the MNIST illustration used in the manuscript. For each digit, we work with a sample of 500 images and the corresponding within-digit pairwise Wasserstein distances. The goal is to estimate within-class heterogeneity, compare the digit classes, and identify the images with the largest empirical eccentricities.

```{r setup}
rm(list = ls())

library(ggplot2)
library(patchwork)
library(grid)

source("src.R")
```

### Run the digit-wise analysis

We first compute the proposed heterogeneity summaries for the ten digit classes. The analysis uses the squared transform (\psi(t)=t^2), along with the corresponding standard errors and Wald intervals.

```{r analyze-mnist}
results <- analyse_mnist_digits(
  data_dir = file.path(getwd(), "data", "distances"),
  digits = 0:9,
  distance_is_squared = FALSE,
  main_transform = "squared",
  do_bounded_sensitivity = TRUE,
  bounded_c0_rule = "global_median",
  alpha = 0.05,
  D0 = 0,
  top_k = 12
)
```

### Within-digit heterogeneity table

The main output is the within-digit summary table. It reports the heterogeneity estimate, its standard error, the 95% confidence interval, and the induced rank across digits.

```{r summary-table}
results$summary
```

For convenience, the digits ranked from most to least heterogeneous are shown below.
```{r ordered-summary}
results$summary[order(-results$summary$estimate), ]
```
These results should reproduce the main pattern reported in the manuscript: digits such as 5, 2, and 7 are more heterogeneous, whereas digits such as 0 and 1 are comparatively stable

### Pairwise comparisons

We also compare digit classes pairwise using the two-sample statistic.
```{r pairwise}
results$pairwise[order(results$pairwise$p_value_bh), ]
```
The smallest adjusted $p$-value correspond to pairs of digits with the most distinct levels of heterogeneity.

### Eccentric observations

We extract the most eccentric observations for each digit.

```{r top-eccentric}
results$top_eccentric
```

For illustration, we focus on the three most heterogeneous digits.
```{r select-top-digits}
top_digits <- as.character(
  results$summary$digit[order(-results$summary$estimate)][1:3]
)

subset(results$top_eccentric, digit %in% top_digits)
```

### Visualizing eccentric images

We now display the most eccentric images from the three digit classes with the largest heterogeneity estimates. We use the empirical eccentricities computed above and extract the top five images within each of the three most heterogeneous digits.
```{r load-images}
load(file.path(getwd(), "data", "per-digit-500.RData"))

top_digits <- as.character(
  results$summary$digit[order(-results$summary$estimate)][1:3]
)

top_eccentric_subset <- do.call(
  rbind,
  lapply(top_digits, function(d) {
    df <- subset(results$top_eccentric, digit == d)
    head(df[order(-df$ghat, df$obs_index), ], 5)
  })
)

top_eccentric_subset
```

```{r build-plot-data}
image_to_df <- function(mat, row_group, col_id) {
  mat <- t(mat[, nrow(mat):1])  # rotate 90 degrees counterclockwise

  nr <- nrow(mat)
  nc <- ncol(mat)

  df <- expand.grid(
    y = seq_len(nr),
    x = seq_len(nc)
  )

  df$y_plot <- nr - df$y + 1
  df$intensity <- as.vector(mat)
  df$row_group <- row_group
  df$col_id <- col_id

  df
}

plot_df_list <- lapply(seq_along(top_digits), function(k) {
  d <- top_digits[k]
  idx <- top_eccentric_subset$obs_index[top_eccentric_subset$digit == d]
  imgs <- list_images[[as.numeric(d) + 1]][idx]

  do.call(
    rbind,
    Map(
      function(mat, j) image_to_df(mat, row_group = paste0("digit", d), col_id = j),
      imgs,
      seq_along(imgs)
    )
  )
})

plot_df <- do.call(rbind, plot_df_list)

plot_df$row_group <- factor(
  plot_df$row_group,
  levels = paste0("digit", top_digits)
)
plot_df$col_id <- factor(plot_df$col_id, levels = 1:5)
```

```{r build-label-data}
label_df <- do.call(
  rbind,
  lapply(top_digits, function(d) {
    df <- subset(top_eccentric_subset, digit == d)
    data.frame(
      row_group = paste0("digit", d),
      col_id = factor(seq_len(nrow(df)), levels = 1:5),
      x = 14.5,
      y_plot = 27,
      label = sprintf("hat(g)[i] == %.2f", df$ghat),
      stringsAsFactors = FALSE
    )
  })
)

label_df$row_group <- factor(
  label_df$row_group,
  levels = paste0("digit", top_digits)
)
label_df$col_id <- factor(label_df$col_id, levels = 1:5)
```

```{r plot-images, fig.align="center", out.width="90%"}
ggplot(plot_df, aes(x = x, y = y_plot, fill = intensity)) +
  geom_raster() +
  scale_fill_gradient(low = "white", high = "black", guide = "none") +
  coord_equal() +
  facet_grid(row_group ~ col_id) +
  geom_label(
    data = label_df,
    aes(x = x, y = y_plot, label = label),
    inherit.aes = FALSE,
    parse = TRUE,
    size = 2.6,
    fill = "white",
    alpha = 0.8,
    label.size = 0,
    label.padding = unit(0.08, "lines"),
    colour = "black"
  ) +
  theme_void() +
  theme(
    strip.text = element_blank(),
    panel.spacing = unit(0, "lines"),
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.4
    )
  )
```

The rows correspond to the three most heterogeneous digit classes, ordered from largest to smaller estimated heterogeneity. Within each row, the images are ordered by decreasing empirical eccentricity. This makes the figure fully reproducible and directly aligned with the proposed interpretation of $\hat g_i$ as an observation-level contribution to within-class heterogeneity.

### Summary

This example shows that the proposed framework provides both:

- a **global comparison** of digit classes through heterogeneity estimates, and 
- a **local explanation** via empirical eccentricities that identify influential observations.